---
title: "OSAA"
author: "Mengyu Li"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(maps)
library(mapdata)
suppressMessages(library("tidyverse"))
library(ggmap)
library(DT)
library(geosphere)
library(here)
library (pander, quietly = TRUE)
library (magrittr, quietly = TRUE) #Used for pipes/dplyr
library(dplyr)
library (ROI, quietly = TRUE)
library (ROI.plugin.glpk, quietly = TRUE)
library (ROI.plugin.symphony, quietly = TRUE)
library (ompr, quietly = TRUE)
library (ompr.roi, quietly = TRUE)
library(readxl)
library(plotly)
```

```{r}


##PART 2 - LOAD THE REQUIRED DATASET AND GET GEOCODING FOR ALL SCHOOLS
Football <- read_excel("OSAA/Prj.xls")
pander(head(Football))

##At this point we can start creating the matrix of distances to figure out the new leagues
# Setting locations as one, rather than separate columns
Football$location <- paste0(Football$Address, ",", Football$City, ", OR ", Football$'Zip Code')
Football$location

# This function geocodes a location (find latitude and longitude) using the Data Science toolkit
geo <- geocode(location = Football$location, output="latlon", source="dsk")

# pasting new lon/lat info to Football table
Football$lon <- geo$lon
Football$lat <- geo$lat

# making a smaller table w/ just longitude and latitude to figure out distances
geo_school <- cbind(Football$lon, Football$lat)
colnames(geo_school) <- c("Longitude", "Latitude")

# lets take a look at the table of distances to make sure all is correct, but we don't want to
# put any extra columns into the geo_school table, so we'll just create a new table that we 
# save as csv 
geo_school_csv <- geo_school
rownames(geo_school_csv) <- Football$School

#Write new data to csv so we can work from that
write.csv(geo_school_csv, file = "lat_lon_6A.csv")

#read the csv into the program
geo_school_csv <- read_csv("lat_lon_6A.csv")
#View(geo_school_csv)


###PART 2.1 - CALCULATE TOTAL MILEAGE OF CURRENT OSAA LEAGUES TRAVEL
##First, we'll make separate tables for each of the leagues within the 6A classification
sixA_lat_lon <- cbind.data.frame(Football$School, geo_school, Football$`Regular League`)
#Here we make a dataframe w/ the school names, their latitude/longitude, and their listed regular league.

sixA_L1 <- subset(sixA_lat_lon, Football$`Regular League` =="1 Portland Interscholastic League")
#Here we make a subset of the table above, so that we only see schools within a specific league
#First, we'll look at League 1, which is the Portland Interscholastic League

sixA_L1_geo <- cbind.data.frame(sixA_L1$Longitude, sixA_L1$Latitude)
#Here a make a new table w/ only the longitude and latitude of the schools within the league selected above
#Specifically,  League 1, which is the Portland Interscholastic League
#I have to make this new table because the distGeo function below will only work with a table w/ 2 columns

dist_L1 <- distm(sixA_L1_geo, fun = distGeo)/1609.344
# Here we create the distance matrix for schools within the Portland Interscholastic League
#Dividing by 1609 to convert from meters to miles 

sum(dist_L1)
#Add up all the distances to get total mileage for schools within the Portland Interscholastic League

##Then we'll repeat the above for each of the leagues within the 6A classification

# 2 Metro League
sixA_L2 <- subset(sixA_lat_lon, Football$`Regular League` =="2 Metro League")
sixA_L2_geo <- cbind.data.frame(sixA_L2$Longitude, sixA_L2$Latitude)
dist_L2 <- distm(sixA_L2_geo, fun = distGeo)/1609.344

# 3 Pacific Conference
sixA_L3 <- subset(sixA_lat_lon, Football$`Regular League` =="3 Pacific Conference")
sixA_L3_geo <- cbind.data.frame(sixA_L3$Longitude, sixA_L3$Latitude)
dist_L3 <- distm(sixA_L3_geo, fun = distGeo)/1609.344

# 4 Mt. Hood Conference
sixA_L4 <- subset(sixA_lat_lon, Football$`Regular League` =="4 Mt. Hood Conference")
sixA_L4_geo <- cbind.data.frame(sixA_L4$Longitude, sixA_L4$Latitude)
dist_L4 <- distm(sixA_L4_geo, fun = distGeo)/1609.344

# 5 Three Rivers League
sixA_L5 <- subset(sixA_lat_lon, Football$`Regular League` =="5 Three Rivers League")
sixA_L5_geo <- cbind.data.frame(sixA_L5$Longitude, sixA_L5$Latitude)
dist_L5 <- distm(sixA_L5_geo, fun = distGeo)/1609.344

# 6 Mountain Valley Conference
sixA_L6 <- subset(sixA_lat_lon, Football$`Regular League` =="6 Mountain Valley Conference")
sixA_L6_geo <- cbind.data.frame(sixA_L6$Longitude, sixA_L6$Latitude)
dist_L6 <- distm(sixA_L6_geo, fun = distGeo)/1609.344

# 7 Southwest Conference
sixA_L7 <- subset(sixA_lat_lon, Football$`Regular League` =="7 Southwest Conference")
sixA_L7_geo <- cbind.data.frame(sixA_L7$Longitude, sixA_L7$Latitude)
dist_L7 <- distm(sixA_L7_geo, fun = distGeo)/1609.344

##Now that baseline distances are collected, we'll put all into one neat table
sum_col <- rbind(sum(dist_L1), sum(dist_L2), sum(dist_L3),sum(dist_L4),sum(dist_L5),sum(dist_L6), sum(dist_L7), (sum(dist_L1, dist_L2, dist_L3, dist_L4, dist_L5, dist_L6, dist_L7)))
sum_col_r <- round(sum_col, digits = 2)
L_col <- rbind("1 Portland Interscholastic League", "2 Metro League", "3 Pacific Conference", 
               "4 Mt. Hood Conference", "5 Three Rivers League","6 Mountain Valley Conference", 
               "7 Southwest Conference", "Total 6A Mileage (all leagues)")
Baseline <- cbind(L_col, sum_col_r)
colnames(Baseline) <- c("6A Leagues", "Mileage")
pander(Baseline, caption = "Current OSAA Leagues & Mileage")

### PART 2.2 - DISTANCE MATRIX
#turning the longitude latitude numbers into distances. 
#Dividing by 1609 to convert from meters to miles
distances <- distm(geo_school, fun = distGeo)/1609.344
d <- round(distances, digits=2)

#Connecting the school names to the distance matrix, columns and rows
n <-Football$School
colnames(d) <- n
rownames(d) <- n

#double check on object types (numbers should be #, words strings)
str(d)

dist_matrix <- d

#Write new data to csv so we can work from that
write.csv(d, file = "OSAADistance.csv")

#read the csv into the program
OSAADistance <- read_csv("OSAADistance.csv")
#View(OSAADistance)
```




```{r}

#----------------------------------------------------
# Visualize locations of teams and pre-group teams into three (3) leagues
#---------------------------------------------------

dfGEO <- as.data.frame(geo_school)
rownames(dfGEO) <- n 
colnames(dfGEO) <- c('Longitude', 'Latitude')
gg<- ggplot()
gg <- gg + geom_point(data= dfGEO, aes(x = Longitude, y = Latitude, color ='red'))
gg <- gg + coord_map()
gg
# 1st league
l1 <- subset(dfGEO, Latitude < 44.5)
l1
# identify 2nd league
gg <- NULL
gg<- ggplot()
gg <- gg + geom_point(data= filter(dfGEO, Latitude > 44.5), aes(x = Longitude, y = Latitude, color ='red'))
gg <- gg + coord_map()
ggplotly(gg)
# 2nd league
l2<-subset(dfGEO, Latitude <= 45.21869 & Latitude > 44.5)
l2
# Identify 3rd league
l3<-subset(dfGEO, Latitude > 44.5 & Longitude >= -122.54)
l3
# Remaining Team Names
rtn<-setdiff(setdiff(setdiff(rownames(dfGEO), rownames(l1)), rownames(l2)), rownames(l3))
rtn

#------------------------------------
# Optimization models
#------------------------------------
# extract dist_matrix for the remaining teams of optimization
dim(dist_matrix)
dist_matrix_rt <- dist_matrix[rtn,rtn]
mDist <- dist_matrix_rt
nTeam <- nrow(mDist)
minTeams <- 6
maxTeams <- 8
diag(mDist) <- 0

```


##Formula
Here, we have $X_i_j$, $X_j_k$, $X_i_k$, to represent games between school i & schoolj, school j & school k, school i & school k.$x_i_j$ = 1 means school i and school j are in the same league  $D_i_j$ represents the distance travel between school i & j, our objective is to minimize the total distance to travel within the league. 

$$
 \begin{split}
 \begin{aligned}
    \text{Minimize  }  & \sum_{i} \sum_{j|j>i} D_{i,j} X_{i,j}*2 \\
    \text{subject to }&\sum_{j>i}x_{i,j} + \sum_{k<i}x_{k,i}  >= minTeams-1, \forall i\\
                       & \sum_{j>i}x_{i,j} + \sum_{k<i}x_{k,i}  <= maxTeams-1, \forall i\\
                       & \ 1- x_{i,j} - x_{i,k} + x_{j,k}\geq 0; \forall \; i,j,k \wedge j>i,k>j\\
                       & x_{i,j} \in(0,1)  \;;\forall i,j=1:nTeam\wedge j>i
  \end{aligned}
  \end{split}

$$
The optimizatin objective is to minimize the traveling distance between teams. The One-way trip between two teams can be calculated as $D_i_j$ * $X_i_j$. If the distance of round-trips is interested, the one-way distance needs to be multiplied by 2 to get the total distance. Neverthless, it is supposed to have no impact on the optimal solution.

Given a full matrix (team number * team number) which has binary values to indicate if two teams in the league, the matrix is essentially symmetric by the diagnoa
Considering the symmetric characteristic in the relationship matrix between teams, we only need to 





##The 2nd constraint facilitates the interpretation of the minimum requirements in a league. Since the distance between a team and itself is 0, this constraint actually can help reduce the solution space.

##The 3rd constraint ensures if teams i and team j in the same league AND teams i and team k in the same league, then teams j and k are in the same league. The beauty of the expression lies in its combination of the constraint of symmetry that effectively imposes reciprocal relationship between teams.





```{r}
# ------------------------------------
# Optimization model 
# ------------------------------------
start_time <- Sys.time()
result <- MIPModel() %>%
  add_variable(x[i,j], j>i, i = 1:nTeam, j = 1:nTeam, type="binary") %>%
  set_objective(sum_expr(2 * mDist[i,j]*x[i,j], j>i, i=1:nTeam, j=1:nTeam),"min") %>%
  add_constraint(sum_expr(x[i,j], j>i, j=1:nTeam) + sum_expr(x[k,i], k<i, k=1:nTeam) >= minTeams - 1, i=1:nTeam) %>%
  add_constraint(sum_expr(x[i,j], j>i, j=1:nTeam) + sum_expr(x[k,i], k<i, k=1:nTeam) <= maxTeams - 1, i=1:nTeam) %>%
  add_constraint((1 - x[i, j] - x[i, k]) + x[j, k] >= 0, j>i, k>j, i=1:nTeam, j=1:nTeam, k=1:nTeam) %>%
  solve_model(with_ROI(solver = "symphony"))
end_time <- Sys.time() - start_time
end_time
print(solver_status(result))
print(result)
View(result$solution)

#print(matrix(result$solution,ncol = nTeam))
# extract leagues with teams from the results
lsLeagues <- NULL
mTeams <- matrix(0, nrow =  nrow(mDist), ncol =  nrow(mDist))
fnFormTeamMatchMatrix <- function(mBase, vVals, nameLetter) {
  m <- mBase
  for (i in seq(1 : nrow(mBase)-1)){
    for (j in seq(i+1: ncol(mBase))){
      val <- vVals[paste0(nameLetter, '[', i, ',', j, ']')]
      m[i, j] <- val
    }
  }
  
  for (j in seq(1: nrow(mBase))){
    for (i in seq(1: j)){
      m[j, i] <- m[i, j]
    }
  }
  diag(m) <- 1
  return(m)
}

fnLeaguesWithTeams <- function(m) {
  for (i in seq(1 : nrow(m))){
    lsLeagues[[i]] <- which(m[i,] %in% 1)
  }
  return(lsLeagues)
}

mTeamMatch <- matrix(0, nrow = nrow(mDist), ncol =  nrow(mDist))
mTeamMatch <- fnFormTeamMatchMatrix(mTeamMatch, result$solution, 'x')
View(mTeamMatch)
lsLeagues <- unique(fnLeaguesWithTeams(mTeamMatch))
lsLeagues # teams in each leagues
lIdx <- 4
for (i in lsLeagues){
  print(paste0('League ', lIdx, ': '))
  print(rtn[i])
  lIdx <- lIdx +1
}
```



